Formalization of Forcing in Isabelle/ZF \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2020-03-09 IJCAR2020/rebuttal.txt

We would like to thank the referees for their
careful reading and detailed criticism.

> ----------------------- REVIEW 1 ---------------------
> 
> The paper's main limitation is its treatment of related work. On
> general principles, the main discussion of related work should be near
> the end of the paper, where it could be compared with the work just
> presented. There, it would be appropriate to contrast what was done in
> Paulson's work on constructibility with the current work on
> forcing. More to the point, the recent work by Han and van Doorn on
> forcing in Lean deserves more discussion.

We will gladly incorporate this changes into a revised version.

> ----------------------- REVIEW 2 ---------------------
> 
> In a single
> publication, Han and van Doorn using forcing to show the independence
> of the continuum hypothesis, whereas the authors of this paper are on
> their third publication, and still seem to be at least one paper away
> from showing any independence result.

We believe that the previous assertions are not entirely accurate. As
it is clear from our introduction, our second manuscript on the
subject is not published (and it won't be) in any venue other than the
arXiv. (As a side note, Han & van Doorn have two papers detailing
the full account of their formalization.)

> The main result in this paper is
> simply that forcing extensions satisfy replacement and choice, which
> seem to be fairly straightforward.

In some sense, this is a misstatement of our achievements. The "Main
theorem" of our paper is just the icing on the cake, since there is no
other way to prove this result other than formalizing the machinery of
forcing, which is the real achievement of this paper. Han & van Doorn
[CPP 2020] hint that the formalization of forcing in set theoretical
foundations is practically impossible.

> 
> 
> ----------------------- REVIEW 3 ---------------------
> This work presents a formalization of forcing in Isabelle/ZF and the 
> proof of a first lemma about the forcing relation.

We would like to point out that we proved the fundamental theorems of
forcing, which are the main tools to work with the forcing relation.

### 2020-03-09 IJCAR2020/rebuttal-miguel.txt

We would like to thank the referees for their
careful reading and detailed criticism.

> ----------------------- REVIEW 1 ---------------------
>
> The paper's main limitation is its treatment of related work. On
> general principles, the main discussion of related work should be near
> the end of the paper, where it could be compared with the work just
> presented. There, it would be appropriate to contrast what was done in
> Paulson's work on constructibility with the current work on
> forcing. More to the point, the recent work by Han and van Doorn on
> forcing in Lean deserves more discussion.

We will gladly incorporate this changes into a revised version.

> ----------------------- REVIEW 2 ---------------------
>
> In a single publication, Han and van Doorn using forcing to show the
> independence of the continuum hypothesis, whereas the authors of
> this paper are on their third publication, and still seem to be at
> least one paper away from showing any independence result.

We knew of Han and van Doorn when we were somewhat advanced in our
project and decided to continue because of the differences: a
different meta-theory and a different approach to forcing. We think
that this paper is interesting per se (we can agree that the arXiv
paper is a bit premature) as it establish a ground on which do forcing
arguments.

> The main result in this paper is simply that forcing extensions
> satisfy replacement and choice, which seem to be fairly
> straightforward.
> ----------------------- REVIEW 3 ---------------------
> This work presents a formalization of forcing in Isabelle/ZF and the
> proof of a first lemma about the forcing relation.

We would like to point that the main result may seem straightforward
thanks to the more substantial contributin of the paper: the
definition of the forces function and the proof of the forcing
theorems. Han & van Doorn [CPP 2020] hint that the formalization of
forcing using the ctm and in a set theoretical foundations is more
intricate than via the Boolean-valued models on type-theory.

### 2020-03-09 IJCAR2020/rebuttal-miguel.txt

> ----------------------- REVIEW 1 ---------------------
>
> The paper's main limitation is its treatment of related work. On
> general principles, the main discussion of related work should be near
> the end of the paper, where it could be compared with the work just
> presented. There, it would be appropriate to contrast what was done in
> Paulson's work on constructibility with the current work on
> forcing. More to the point, the recent work by Han and van Doorn on
> forcing in Lean deserves more discussion.

We will gladly incorporate this changes into a revised version.

> ----------------------- REVIEW 2 ---------------------
>
> In a single publication, Han and van Doorn using forcing to show the
> independence of the continuum hypothesis, whereas the authors of
> this paper are on their third publication, and still seem to be at
> least one paper away from showing any independence result.

We knew of Han and van Doorn when we were somewhat advanced in our
project and decided to continue because of the differences: a
different meta-theory and a different approach to forcing. We think
that this paper is interesting per se (we can agree that the arXiv
paper is a bit premature) as it establish a ground on which to do
forcing arguments.

### 2020-03-12 IJCAR2020/rebuttal.txt

We would like to thank the referees for their
careful reading and detailed criticism.

> ----------------------- REVIEW 1 ---------------------
> 
> The paper's main limitation is its treatment of related work. On
> general principles, the main discussion of related work should be near
> the end of the paper, where it could be compared with the work just
> presented. There, it would be appropriate to contrast what was done in
> Paulson's work on constructibility with the current work on
> forcing. More to the point, the recent work by Han and van Doorn on
> forcing in Lean deserves more discussion.

We will gladly incorporate these changes and discussion on the
questions posed into a revised version.

> ----------------------- REVIEW 2 ---------------------
> 
> In a single
> publication, Han and van Doorn using forcing to show the independence
> of the continuum hypothesis, whereas the authors of this paper are on
> their third publication, and still seem to be at least one paper away
> from showing any independence result. In particular, CH will require
> some careful cardinal combinatorics.

We knew of Han and van Doorn when we were somewhat advanced in our
project and decided to continue because of the differences:

 1) a weaker meta-theory (Han & van Doorn [CPP 2020] hint that the
 formalization of forcing in set theoretical foundations is practically
 impossible); and

 2) an approach to forcing (ctm) used in a important fraction of the
 literature.

We agree that the remaining work towards the independence of CH will
require one more paper. Notice that, in fact, Han & van Doorn have two
papers (ITP2019, CPP2020) detailing the full account of their
formalization. (As a side note, our second manuscript on the subject
is not published.)

@@

> The main result in this paper is
> simply that forcing extensions satisfy replacement and choice, which
> seem to be fairly straightforward.

This is a misstatement of our achievements. Our "Main Theorem" may
seem straightforward thanks to the more substantial contribution of
the paper: the definition of the forcing relation and the proof of the
forcing theorems. The Main Theorem is a succinct way to demonstrate
the forcing machinery, since there is no other way to produce
extensions of ctms.

> ----------------------- REVIEW 3 ---------------------
> This work presents a formalization of forcing in Isabelle/ZF and the 
> proof of a first lemma about the forcing relation.

### 2020-03-13 IJCAR2020/rebuttal-miguel.txt

> ----------------------- REVIEW 2 ---------------------
>
> In a single publication, Han and van Doorn using forcing to show the
> independence of the continuum hypothesis, whereas the authors of
> this paper are on their third publication, and still seem to be at
> least one paper away from showing any independence result.

We knew of Han and van Doorn when we were somewhat advanced in our
project and decided to continue because of the differences: a
different meta-theory and a different approach to forcing. We think
that this paper is interesting per se (we can agree that the arXiv
paper is a bit premature) as it establish a ground on which to do
forcing arguments.

> The main result in this paper is simply that forcing extensions
> satisfy replacement and choice, which seem to be fairly
> straightforward.
> ----------------------- REVIEW 3 ---------------------
> This work presents a formalization of forcing in Isabelle/ZF and the
> proof of a first lemma about the forcing relation.

We would like to point that the main result may seem straightforward
thanks to the more substantial contribution of the paper: the
definition of the forces function and the proof of the forcing
theorems. Han & van Doorn [CPP 2020] hint that the formalization of
forcing using the ctm and in a set theoretical foundations is more
intricate than via the Boolean-valued models on type-theory.

### 2020-03-13 IJCAR2020/absoluteness.tex

Apart from the axiom schemes, the $\ZFC$ 
axioms are initially stated as predicates on classes (that is,
of type \isatt{(i{\isasymRightarrow}o){\isasymRightarrow}o}); some
of them were already defined by Paulson, and this formulation allows a
better interaction with \isatt{ZF-Constructible}. 
The axioms of Pairing, Union, Foundation,
Extensionality, and Infinity are relativizations of the respective
traditional 
first-order sentences to the class argument. For the Axiom of Choice
we selected a version best suited for the work with transitive
models: the relativization of a sentence stating that for every $x$
there is surjection from an ordinal onto $x$. Finally, Separation and
Replacement were treated separately to effectively obtain first-order
versions afterwards. It is to be noted that predicates in Isabelle/ZF
are akin to second order variables and thus do not correspond to first-order
formulas.
%% in particular, there is no induction principle for functions
%% of type \isatt{i{\isasymRightarrow}o}. 
For that reason, Separation and Replacement predicate \emph{on the
  satisfaction} of a formula $\phi$.  We improved their specification,
with respect to our previous \cite{2019arXiv190103313G}, by lifting
the arity restriction for the parameter $\phi$; consequently we
get rid of tupling and thus various proofs are now slicker.

>>>>

Apart from the axiom schemes, the $\ZFC$ axioms are initially stated
as predicates on classes (that is, of type
\isatt{(i{\isasymRightarrow}o){\isasymRightarrow}o}); this formulation
allows a better interaction with \isatt{ZF-Constructible}.  The axioms
of Pairing, Union, Foundation, Extensionality, and Infinity are
relativizations of the respective traditional first-order sentences to
the class argument. For the Axiom of Choice we selected a version best
suited for the work with transitive models: the relativization of a
sentence stating that for every $x$ there is surjection from an
ordinal onto $x$. Finally, Separation and Replacement were treated
separately to effectively obtain first-order versions afterwards. It
is to be noted that predicates in Isabelle/ZF are akin to second order
variables and thus do not correspond to first-order formulas.
%% in particular, there is no induction principle for functions
%% of type \isatt{i{\isasymRightarrow}o}. 
For that reason, Separation and Replacement predicate \emph{on the
  satisfaction} of a formula $\phi$.  We improved their specification,
with respect to our previous \cite{2019arXiv190103313G}, by lifting
the arity restriction for the parameter $\phi$; consequently we
get rid of tupling and thus various proofs are now slicker.

### 2020-03-13 IJCAR2020/forces_definition.tex

In fact, given a forcing notion (ie. a poset with a top element) Kunen
defines meta-theoretically the forcing relation $p \Vdash \phi$ as the
satisfaction by $M$ of $\forceisa(\phi)$ in an environment containing
the forcing notion and $p$; next he shows that this relation is
definable. To make sense of this in our formalization, we started with
an internalized relation and then proved the equivalence of that
relation with the semantic version of the forcing relation.  The
latter relation is captured by the following abbreviation, after
defining $\forceisa$:
\begin{isabelle}
\ \ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ env\ {\isasymequiv}\ sats{\isacharparenleft}M{\isacharcomma}forces{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isacharcomma}\ {\isacharbrackleft}p{\isacharcomma}P{\isacharcomma}leq{\isacharcomma}one{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

>>>>

In fact, given a forcing notion $\PP$ (i.e. a preorder with a top element)
in a ctm $M$,
Kunen defines the \emph{forcing relation} model-theoretically 
by considering all extensions $M[G]$ with $G$ generic for $\PP$.
Then two fundamental results are proved, the Truth Lemma and the
Definability Lemma; but the proof of the first one is based on the
formula that witnesses Definability. To make sense of this in our 
formalization, we started with the internalized relation and then
proved that it is equivalent to the semantic version 
(``\isatt{definition{\isacharunderscore}of{\isacharunderscore}forces},'' in
the next section).
For that reason, the usual notation of the forcing relation 
$p \Vdash \phi\ \mathit{env}$ (for $\mathit{env}$ a list of
``names''), abbreviates in our code the
satisfaction by $M$ of $\forceisa(\phi)$:
\begin{isabelle}
\ \ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ env\ \ \ {\isasymequiv}\ \ \ M{\isacharcomma}\ {\isacharparenleft}{\isacharbrackleft}p{\isacharcomma}P{\isacharcomma}leq{\isacharcomma}one{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}
    {\isasymTurnstile}\ forces{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
